Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-04-13 ThesisJP/Paper2/cos.tex
Fixed figure layout. Checked all labels.

In the next section we present background information on Capon beamforming in both element-space and beamspace. We also give an introduction to the GPU computing model. In Section \ref{sec:meth} we give a summary of our parallel implementation of the element-space Capon beamformer, before the GPU implementation of the beamspace Capon beamformer is presented in Section \ref{sec:bs}. Benchmark results and resulting images for both implementations are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}. Finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=0.5\textwidth]{gfx/benchmark_bar_64_80_540.png} \label{fig:benchUS64}}\\
\hfil
\subfloat[]{\includegraphics[width=0.5\textwidth]{gfx/benchmark_bar_96_120_540.png}%
\label{fig:benchUS96}}}
%\centerline{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_540.png}}
\caption{Benchmarks of GPU Capon beamforming in element space for a cardiac image covering a $70\degree$ wide $15.4$ cm deep sector. Execution time is plotted as a function of subarray length $L$. The figures show execution times for the three steps listed in Section \ref{sec:meth}: Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.}
\label{fig:bench}
\end{figure*}

\begin{figure*}[!t]
\centerline{\subfloat[]{\includegraphics[width=0.5\textwidth]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}
\hfil
\subfloat[]{\includegraphics[width=0.5\textwidth]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
\label{fig:benchBS96}}}
\caption{Benchmarks of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15.4$ cm deep sector. Execution time is plotted as a function of subarray length $L$.  If $L$ had been fixed at $M/2$ and execution time was plotted against $N_b$, the solver would have shown the same quadratic growth as in Fig.\ \ref{fig:bench}. When $N_b$ is fixed, calculating the covariance matrix and the BS-transform is inverse proportional and proportional to $L$ respectively. The large jumps in execution times are caused by adaptive kernel launch configurations, which depends on $M$, $L$ and $N_b$.}
\label{fig:benchBS}
\end{figure*}
In this section we present benchmarks of the implementations presented in Sections \ref{sec:meth} and \ref{sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision performance, 6 GB global memory, and 144 GB/s memory bandwidth). Problem sizes are calculated based on 64 and 96 element phased arrays with a 3.4 MHz center frequency. Assuming a system bandwidth of 80\% (relative to the center frequency) we need a complex sampling frequency of 2.7 MHz, and when imaging down to a depth of 15.4 cm we have approximately 540 samples per range line. The angular sampling is given by the Rayleigh criterion divided by two \cite{Hergum2007}, 
\begin{equation}\label{eq:rayleigh}
\delta{}\theta = \lambda/(2Mp),
\end{equation}
where $\lambda$ is the pulse wavelength and $p$ is the element pitch. 
To sample a $70\degree$ sector with $64$- and $96$-element arrays with $p=\lambda/2$ we therefore need approximately 80 and 120 receive beams respectively.

@@

\begin{figure*}[!t]
\centerline{
\subfloat[]{\includegraphics[width=0.3\textwidth]{gfx/das_invivo.png} \label{fig:invivoDAS}}
\hfil
\subfloat[]{\includegraphics[width=0.3\textwidth]{gfx/capon_invivo.png} \label{fig:invivoMV}}
\hfil
\subfloat[]{
\includegraphics[width=0.3\textwidth]{gfx/capon_bs_invivo.png} \label{fig:invivoBS}}
\hfil
\includegraphics[height=0.3\textwidth]{gfx/colorBarInvivo2.png}
}
\caption{\textit{In-vivo} harmonic cardiac ultrasound image acquired using a 3.5 MHz, 64 elements phased array. Number of receive beams for the $70\degree$ sector is 98, and number of samples in range is 610. Dynamic range is -20 dB to 20 dB with mean value represented as 0 dB. The image is the first frame in the attached videos. a) Delay-and-sum with uniform weights \multimedia{Media-Movie 1}. b) Element-space Capon ($L=32, K=2, d=0.1$) \multimedia{Media-Movie 2}. c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 3}.}
\label{fig:invivo}
\end{figure*}

@@

\section{Discussion}\label{sec:dis}
In this paper we have presented a parallel implementation of the BS-Capon beamformer for high-resolution cardiac ultrasound imaging. The results show that it is now possible to apply BS-Capon in real-time in cardiac ultrasound imaging by utilizing the power of a GPU. According to Fig.\ \ref{fig:benchBS} it is also possible to process two parallel receive lines per transmit beam (for both $M=64$ and $M=96$. $L=M/2$). %However, since we had to use oversampling in our simulation, these extra lines might be needed for increased sampling instead of increasing the frame rate. If parallel receive lines are used to obtain denser sampling, the acquisition time remains the same but the processing time is increased with a factor equal to the amount of oversampling. If parallel receive lines are used to increase the frame rate, the acquisition time is lowered, and the processing time remains constant. Hence, four parallel receive lines are not possible to process in real-time using BS-Capon with $N_b=3$ in combination with our single GPU setup.

@@

\section{Conclusion}\label{sec:con}
In this paper a parallel formulation and implementation of the beamspace Capon beamformer have been presented, and
the beamformer has successfully been implemented on the GPU. For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging was shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters. Image quality was investigated in an \textit{in-vivo} cardiac dataset. These results showed that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution both in element- and beamspace.

>>>>

In the next section we present background information on Capon beamforming in both element-space and beamspace. We also give an introduction to the GPU computing model. In Section \ref{sec:meth} we give a summary of our parallel implementation of the element-space Capon beamformer, before the GPU implementation of the beamspace Capon beamformer is presented in Section \ref{sec:bs}. Benchmark results and resulting images for both implementations are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}. Finally we discuss the results in Section \ref{sec:dis_paper2}, and draw conclusions in Section \ref{sec:con_paper2}.

@@

\begin{figure*}[!t]
\centerline{
\begin{tabular}{c}
\subfloat[]{\includegraphics[width=0.7\textwidth]{gfx/benchmark_bar_64_80_540.png} \label{fig:benchUS64}}\\
\subfloat[]{\includegraphics[width=0.7\textwidth]{gfx/benchmark_bar_96_120_540.png}%
\label{fig:benchUS96}}
\end{tabular}
}
\caption{Benchmarks of GPU Capon beamforming in element space for a cardiac image covering a $70\degree$ wide $15.4$ cm deep sector. Execution time is plotted as a function of subarray length $L$. The figures show execution times for the three steps listed in Section \ref{sec:meth}: Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.}
\label{fig:bench}
\end{figure*}

\begin{figure*}[!t]
\centerline{
\begin{tabular}{c}
\subfloat[]{\includegraphics[width=0.7\textwidth]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{fig:benchBS64}}\\
\subfloat[]{\includegraphics[width=0.7\textwidth]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
\label{fig:benchBS96}}
\end{tabular}
}
\caption{Benchmarks of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15.4$ cm deep sector. Execution time is plotted as a function of subarray length $L$.  If $L$ had been fixed at $M/2$ and execution time was plotted against $N_b$, the solver would have shown the same quadratic growth as in Fig.\ \ref{fig:bench}. When $N_b$ is fixed, calculating the covariance matrix and the BS-transform is inverse proportional and proportional to $L$ respectively. The large jumps in execution times are caused by adaptive kernel launch configurations, which depends on $M$, $L$ and $N_b$.}
\label{fig:benchBS}
\end{figure*}
In this section we present benchmarks of the implementations presented in Sections \ref{sec:meth} and \ref{sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision performance, 6 GB global memory, and 144 GB/s memory bandwidth). Problem sizes are calculated based on 64 and 96 element phased arrays with a 3.4 MHz center frequency. Assuming a system bandwidth of 80\% (relative to the center frequency) we need a complex sampling frequency of 2.7 MHz, and when imaging down to a depth of 15.4 cm we have approximately 540 samples per range line. The angular sampling is given by the Rayleigh criterion divided by two \cite{Hergum2007}, 
\begin{equation}\label{eq:rayleigh}
\delta{}\theta = \lambda/(2Mp),
\end{equation}
where $\lambda$ is the pulse wavelength and $p$ is the element pitch. 
To sample a $70\degree$ sector with $64$- and $96$-element arrays with $p=\lambda/2$ we therefore need approximately 80 and 120 receive beams respectively.

@@

\begin{figure*}[!t]
\centerline{
\begin{tabular}{cc}
\subfloat[]{\includegraphics[width=0.45\textwidth]{gfx/das_invivo.png} \label{fig:invivoDAS}} &
\includegraphics[height=0.45\textwidth]{gfx/colorBarInvivo2.png}\\
\subfloat[]{\includegraphics[width=0.45\textwidth]{gfx/capon_invivo.png} \label{fig:invivoMV}} &
\subfloat[]{\includegraphics[width=0.45\textwidth]{gfx/capon_bs_invivo.png} \label{fig:invivoBS}}
\end{tabular}
}
\caption{\textit{In-vivo} harmonic cardiac ultrasound image acquired using a 3.5 MHz, 64 elements phased array. Number of receive beams for the $70\degree$ sector is 98, and number of samples in range is 610. Dynamic range is -20 dB to 20 dB with mean value represented as 0 dB. The image is the first frame in the attached videos. a) Delay-and-sum with uniform weights \multimedia{Media-Movie 1}. b) Element-space Capon ($L=32, K=2, d=0.1$) \multimedia{Media-Movie 2}. c) Beamspace Capon ($L=32, K=2, N_b=3, d=0.01$) \multimedia{Media-Movie 3}.}
\label{fig:invivo}
\end{figure*}

@@

\section{Discussion}\label{sec:dis_paper2}
In this paper we have presented a parallel implementation of the BS-Capon beamformer for high-resolution cardiac ultrasound imaging. The results show that it is now possible to apply BS-Capon in real-time in cardiac ultrasound imaging by utilizing the power of a GPU. According to Fig.\ \ref{fig:benchBS} it is also possible to process two parallel receive lines per transmit beam (for both $M=64$ and $M=96$. $L=M/2$). %However, since we had to use oversampling in our simulation, these extra lines might be needed for increased sampling instead of increasing the frame rate. If parallel receive lines are used to obtain denser sampling, the acquisition time remains the same but the processing time is increased with a factor equal to the amount of oversampling. If parallel receive lines are used to increase the frame rate, the acquisition time is lowered, and the processing time remains constant. Hence, four parallel receive lines are not possible to process in real-time using BS-Capon with $N_b=3$ in combination with our single GPU setup.

@@

\section{Conclusion}\label{sec:con_paper2}
In this paper a parallel formulation and implementation of the beamspace Capon beamformer have been presented, and
the beamformer has successfully been implemented on the GPU. For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging was shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters. Image quality was investigated in an \textit{in-vivo} cardiac dataset. These results showed that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution both in element- and beamspace.
